---
title: "longread"
format: revealjs
editor: visual
---


```{r}

# Load counts file
counts <- read.delim("/Users/sama/Desktop/research/nanopore/gtex_counts_transcript.txt", header = TRUE, stringsAsFactors = FALSE)
gtex_annot <- read_excel("/Users/sama/Desktop/research/nanopore/mtDNA/41586_2022_5035_MOESM4_ESM.xlsx")
```

```{r}

# Clean column names: remove FAK...suffix and replace dots with dashes
clean_names <- names(counts)[-c(1,2)] %>%
  str_remove("\\.FAK.*$") %>%
  str_replace_all("\\.", "-")

# Replace original column names with cleaned ones
names(counts)[-c(1,2)] <- clean_names

# Create a lookup table from sample_id to tissue
annot_lookup <- gtex_annot %>%
  dplyr::select(sample_id, tissue) %>%
  filter(!is.na(tissue)) %>%
  distinct()

# Replace sample IDs in counts with tissue names where matches exist
name_mapping <- tibble(sample_id = names(counts)[-c(1,2)]) %>%
  left_join(annot_lookup, by = "sample_id") %>%
  mutate(final_name = if_else(!is.na(tissue), tissue, sample_id))
names(counts)[-c(1,2)] <- name_mapping$final_name

counts

```

Read counts were averaged across samples from common tissues to obtain representative transcript expression profiles for each tissue type.

```{r}

# Gather tissue columns into long format (excluding TXNAME and GENEID)
long_counts <- counts %>%
  pivot_longer(
    cols = -c(TXNAME, GENEID),  
    names_to = "tissue",
    values_to = "count"
  )

# Average counts for replicates per transcript (TXNAME) and tissue
avg_counts <- long_counts %>%
  group_by(TXNAME, GENEID, tissue) %>%
  summarize(avg_count = mean(count), .groups = "drop")

# Pivot back to wide format (one column per tissue)
counts_avg <- avg_counts %>%
  pivot_wider(
    names_from = tissue,
    values_from = avg_count
  )

```

The GTF annotation file was imported and filtered to retain only exon features. Transcript lengths were then calculated by summing the widths of all exons belonging to each transcript.

```{r}

# Load GTF
gtf <- import("/Users/sama/Desktop/research/nanopore/extended_annotations.gtf")

# Subset GTF to exons
gtf_exons <- gtf[gtf$type == "exon"]

# Compute transcript lengths by summing exon widths
tx_lengths <- gtf_exons %>%
  as.data.frame() %>%
  group_by(transcript_id) %>%
  summarize(length = sum(width)) %>%
  ungroup()

```

TPM values were calculated from the transcript lengths and read counts to enable comparison with short-read RNA-seq TPM data.

```{r}

library(tidyr)

# Rename for clarity
colnames(counts_avg)[1:2] <- c("TXNAME", "GENEID")

# Pivot longer for per-tissue processing
counts_long <- counts_avg %>%
  pivot_longer(cols = -c(TXNAME, GENEID), names_to = "Tissue", values_to = "Count")

# Merge with lengths
counts_long <- counts_long %>%
  left_join(tx_lengths, by = c("TXNAME" = "transcript_id"))

counts_tpm <- counts_long %>%
  mutate(Count_per_kb = Count / (length / 1000)) %>%
  group_by(Tissue) %>%
  mutate(Sum_per_kb = sum(Count_per_kb, na.rm = TRUE),
         TPM = (Count_per_kb / Sum_per_kb) * 1e6) %>%
  ungroup()

tpm_wide <- counts_tpm %>%
  dplyr::select(TXNAME, GENEID, Tissue, TPM) %>%
  tidyr::pivot_wider(names_from = Tissue, values_from = TPM)

tpm_wide_formatted <- tpm_wide %>%
  mutate(across(-c(TXNAME, GENEID), ~ formatC(.x, format = "f", digits = 2)))


```

```{r}

# Load combined transcript list
combined_transcripts <- read.csv("/Users/sama/Desktop/research/combined_transcripts.csv", header = FALSE)
colnames(combined_transcripts) <- c("TXNAME")  

# Filter TPM matrix
long_tpm_filtered <- tpm_wide_formatted %>%
  dplyr::filter(TXNAME %in% combined_transcripts$TXNAME)

```

```{r}

combined_transcripts <- read.csv("/Users/sama/Desktop/research/combined_transcripts.csv", header = FALSE)
colnames(combined_transcripts) <- "transcript_id"
combined_transcripts$transcript_id <- sub("\\..*$", "", combined_transcripts$transcript_id)

short_tpm_filtered <- processed_tpm_avg %>%
  filter(transcript_id %in% combined_transcripts$transcript_id)


```

```{r}

# Get tissue columns 
short_tissues <- colnames(short_tpm_filtered)[!(colnames(short_tpm_filtered) %in% c("transcript_id", "gene_id"))]
long_tissues <- colnames(long_tpm_filtered)[!(colnames(long_tpm_filtered) %in% c("TXNAME", "GENEID"))]

# Find overlaps
common_tissues <- intersect(short_tissues, long_tissues)

# See what overlaps
print(common_tissues)

# Subset both dataframes to just the common tissues + IDs
short_common <- short_tpm_filtered %>%
  dplyr::select(transcript_id, gene_id, all_of(common_tissues))

long_common <- long_tpm_filtered %>%
  dplyr::select(TXNAME, GENEID, all_of(common_tissues))


long_common <- long_common %>%
  rename(transcript_id = TXNAME, gene_id = GENEID)

tpm_comparison <- inner_join(short_common, long_common, by = c("transcript_id", "gene_id"), suffix = c("_short", "_long"))

colnames(tpm_comparison)
```

```{r}

all_cols <- colnames(tpm_comparison)
tissue_names <- gsub("_short|_long", "", all_cols[3:length(all_cols)])  
unique_tissues <- unique(tissue_names)

# Reorder columns to group short/long for each tissue
paired_cols <- c("transcript_id", "gene_id")  # Keep ID columns first

for (tissue in unique_tissues) {
  short_col <- paste0(tissue, "_short")
  long_col <- paste0(tissue, "_long")
  if (short_col %in% all_cols & long_col %in% all_cols) {
    paired_cols <- c(paired_cols, short_col, long_col)
  } else if (short_col %in% all_cols) {
    paired_cols <- c(paired_cols, short_col)
  } else if (long_col %in% all_cols) {
    paired_cols <- c(paired_cols, long_col)
  }
}

# Reorder the dataframe
tpm_comparison_ordered <- tpm_comparison[, paired_cols]

```

```{r}

plot(
  as.numeric(tpm_comparison$`Liver_short`),
  as.numeric(tpm_comparison$`Liver_long`),
  xlab = "Short-read TPM",
  ylab = "Long-read TPM",
  main = "Liver TPM Comparison"
)
abline(0, 1, col = "red", lty = 2)

# Set up a 2x2 plotting layout for 4 plots
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))  # margins: bottom, left, top, right

# Liver
plot(
  as.numeric(tpm_comparison$`Liver_short`),
  as.numeric(tpm_comparison$`Liver_long`),
  xlab = "Short-read TPM",
  ylab = "Long-read TPM",
  main = "Liver TPM Comparison"
)
abline(0, 1, col = "red", lty = 2)

# Brain - Frontal Cortex (BA9)
plot(
  as.numeric(tpm_comparison$`Brain - Frontal Cortex (BA9)_short`),
  as.numeric(tpm_comparison$`Brain - Frontal Cortex (BA9)_long`),
  xlab = "Short-read TPM",
  ylab = "Long-read TPM",
  main = "Brain - Frontal Cortex (BA9) TPM Comparison"
)
abline(0, 1, col = "red", lty = 2)

# Lung
plot(
  as.numeric(tpm_comparison$`Lung_short`),
  as.numeric(tpm_comparison$`Lung_long`),
  xlab = "Short-read TPM",
  ylab = "Long-read TPM",
  main = "Lung TPM Comparison"
)
abline(0, 1, col = "red", lty = 2)

# Heart - Left Ventricle
plot(
  as.numeric(tpm_comparison$`Heart - Left Ventricle_short`),
  as.numeric(tpm_comparison$`Heart - Left Ventricle_long`),
  xlab = "Short-read TPM",
  ylab = "Long-read TPM",
  main = "Heart - Left Ventricle TPM Comparison"
)
abline(0, 1, col = "red", lty = 2)
par(mfrow = c(1,1))



```

### Switching and Dominant transcripts per tissues.

```{r}

# Load necessary library
library(dplyr)

# Remove rows where GENEID or TXNAME starts with "BambuGene"
tpm_cleaned <- tpm_wide_formatted %>%
  filter(!grepl("^BambuGene", GENEID) & !grepl("^BambuTx", TXNAME))

# Identify tissue columns
tissue_cols_long <- names(tpm_cleaned)[3:ncol(tpm_cleaned)]
tpm_cleaned[tissue_cols_long] <- lapply(tpm_cleaned[tissue_cols_long], as.numeric)

# Convert TPM values to percentages per gene per tissue
long_filtered_percentages <- tpm_cleaned %>%
  group_by(GENEID) %>%
  mutate(across(all_of(tissue_cols_long), ~ . / sum(., na.rm = TRUE) * 100)) %>%
  ungroup()

# Select dominant transcript per gene per tissue (≥% expression)
dominant_transcripts_long <- long_filtered_percentages %>%
  group_by(GENEID) %>%
  summarise(across(all_of(tissue_cols_long), 
                   ~ ifelse(any(. >= 60), TXNAME[which.max(.)], "None")))

# Convert "None" values to NA
dominant_transcripts_long <- dominant_transcripts_long %>%
  mutate(across(where(is.character), ~ na_if(.x, "None")))

# Initialise a data frame to collect switching genes
all_switching_genes_long <- data.frame()

# Loop through each gene to check for multiple unique dominant transcripts
for (i in 1:nrow(dominant_transcripts_long)) {
  transcripts <- na.omit(as.character(dominant_transcripts_long[i, -1]))  # Exclude GENEID
  unique_transcripts <- unique(transcripts)
  
  if (length(unique_transcripts) > 1) {
    all_switching_genes_long <- rbind(all_switching_genes_long, dominant_transcripts_long[i, ])
  }
}

all_switching_genes_long


# Get unique gene IDs from high_conf
unique_gene_ids <- unique(mtp_pos$gene_id)

# SFilter all_switching_genes_long for matching GENEID
long_filt_df <- dominant_transcripts_long[dominant_transcripts_long$GENEID %in% unique_gene_ids, ]
filtered_df <- all_switching_genes_long[all_switching_genes_long$GENEID %in% unique_gene_ids, ]




```

```{r}

# Clean column names and convert to character
filtered_df[] <- lapply(filtered_df, as.character)
colnames(filtered_df)[1] <- "gene_id"  # rename for consistency

# Replace NA with "None" on subsetted data
filtered_df_clean <- filtered_df %>%
  mutate(across(-gene_id, ~ ifelse(is.na(.), "None", .)))

# Pivot to long format
plot_data_long <- filtered_df %>%
  pivot_longer(-gene_id, names_to = "Tissue", values_to = "Dominant_Transcript")


```

```{r}


# Select genes to plot using filtered_df$gene_id (lowercase)
genes_to_plot <- filtered_df$gene_id[c(1, 21, 23, 12, 43)]

# Strip version suffixes 
strip_version <- function(x) sub("\\.\\d+$", "", x)
genes_to_plot_clean <- strip_version(genes_to_plot)
long_filtered_percentages$Gene_ID_clean <- strip_version(long_filtered_percentages$Gene_ID)

# Filter long read data using cleaned gene IDs
plot_data <- long_filtered_percentages %>%
  filter(Gene_ID_clean %in% genes_to_plot_clean)

# Pivot longer for plotting
tissue_cols <- setdiff(colnames(long_filtered_percentages), c("Transcript_ID", "Gene_ID", "Gene_ID_clean"))

plot_data_long <- plot_data %>%
  pivot_longer(cols = all_of(tissue_cols), names_to = "Tissue", values_to = "Percent")

# Plot
ggplot(plot_data_long, aes(x = Tissue, y = Percent, fill = Transcript_ID)) +
  geom_bar(stat = "identity", position = "stack", color = "black", size = 0.2) +
  facet_wrap(~ Gene_ID_clean, ncol = 1, scales = "free_y") +
  labs(
    title = "Long-Read Tissue-Specific Dominant Transcript Switching",
    y = "Expression %", x = "Tissue"
  ) +
  scale_fill_discrete(name = "Transcript") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    strip.text = element_text(size = 10, face = "bold"),
    panel.spacing = unit(1, "lines")
  )

```

Check for common switching genes.

```{r}

# Remove version suffix 
switch_gene_ids_clean <- sub("\\..*", "", switch_gene_ids)
all_switching_genes_long <- all_switching_genes_long %>%
  mutate(GENEID_clean = sub("\\..*", "", GENEID))

# Filter to keep only matching genes (ignore version)
common_switching_genes <- all_switching_genes_long %>%
  filter(GENEID_clean %in% switch_gene_ids_clean) %>%
  dplyr::select(-GENEID_clean)  

# View 
common_switching_genes

```

```{r}

# Clean gene IDs 
long_df <- all_switching_genes_long %>%
  mutate(gene_id = sub("\\..*", "", GENEID)) %>%
  dplyr::select(-GENEID)

short_df <- all_switching_genes %>%
  mutate(gene_id = sub("\\..*", "", gene_id))

# Find common genes
common_genes <- intersect(long_df$gene_id, short_df$gene_id)

# Subset to only common genes
long_common <- long_df %>% filter(gene_id %in% common_genes)
short_common <- short_df %>% filter(gene_id %in% common_genes)

# Get common tissues 
tissue_cols_long <- setdiff(colnames(long_common), "gene_id")
tissue_cols_short <- setdiff(colnames(short_common), "gene_id")
common_tissues <- intersect(tissue_cols_long, tissue_cols_short)

# Match and compare transcripts
comparison_df <- data.frame()

for (gene in common_genes) {
  long_row <- long_common %>% filter(gene_id == gene)
  short_row <- short_common %>% filter(gene_id == gene)
  
  for (tissue in common_tissues) {
    long_tx <- as.character(long_row[[tissue]])
    short_tx <- as.character(short_row[[tissue]])
    
    if (is.na(short_tx) || short_tx == "None") {
      status <- NA               # Grey: short no data
    } else if (is.na(long_tx) || long_tx == "None") {
      status <- 2                # Yellow: one has transcript, other no data
    } else if (long_tx == short_tx) {
      status <- 1                # Green: match
    } else {
      status <- 0                # Red: mismatch
    }
    
    comparison_df <- rbind(comparison_df, data.frame(
      gene_id = gene,
      tissue = tissue,
      status = status
    ))
  }
}

heatmap_data <- comparison_df %>%
  pivot_wider(names_from = tissue, values_from = status) %>%
  tibble::column_to_rownames("gene_id")

heatmap_matrix <- as.matrix(heatmap_data)

# Extract metadata from GRanges object
gencode_meta <- as.data.frame(mcols(gencode_transcripts))

# Clean gene_id in gencode_transcripts to remove version numbers
gencode_transcripts_clean <- gencode_meta %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

# Filter to only genes present in heatmap matrix 
gene_map <- gencode_transcripts_clean %>%
  filter(gene_id_clean %in% rownames(heatmap_matrix)) %>%
  dplyr::select(gene_id_clean, gene_name) %>%
  distinct()

# Create a named vector: names = cleaned gene_id, values = gene_name
gene_name_vector <- setNames(gene_map$gene_name, gene_map$gene_id_clean)

# Replace rownames of heatmap_matrix with gene names if available, else keep original gene_id
new_rownames <- sapply(rownames(heatmap_matrix), function(gid) {
  if (gid %in% names(gene_name_vector)) {
    gene_name_vector[gid]
  } else {
    gid  # fallback to gene_id if no match
  }
})

rownames(heatmap_matrix) <- new_rownames

pheatmap(
  heatmap_matrix,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  display_numbers = FALSE,
  color = c("tomato", "lightgreen", "lightyellow"),     # 0=red, 1=green, 2=yellow
  breaks = c(-0.5, 0.5, 1.5, 2.5),
  na_col = "grey",                         # NA short no data
  main = "Transcript Match Between Long-Read and Short-Read"
)

```

```{r}
library(ggplot2)

# Filter out NAs 
status_filtered <- comparison_df$status[!is.na(comparison_df$status)]

# Create a factor with descriptive labels
status_factor <- factor(
  status_filtered,
  levels = c(1, 0, 2),
  labels = c("Match", "No Match", "Expression Discrepancy")
)

# Count and calculate proportions
status_counts_df <- as.data.frame(table(status_factor))
colnames(status_counts_df) <- c("Status", "Count")
total <- sum(status_counts_df$Count)
status_counts_df$Percentage <- (status_counts_df$Count / total) * 100
status_colors <- c("No Match" = "tomato", "Match" = "lightgreen", "Expression Discrepancy" = "lightyellow")

# Plot proportions without grid
ggplot(status_counts_df, aes(x = Status, y = Percentage, fill = Status)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = status_colors) +
  geom_text(aes(label = sprintf("%.1f%%", Percentage)), vjust = -0.5, size = 5) +
  theme_minimal() +
  labs(title = "Proportion of Transcript Match Status",
       x = "Status",
       y = "Percentage (%)") +
  ylim(0, 100) +
  theme(
    text = element_text(size = 14),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

```

